---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

- set global options:
```{r setoptions, echo=TRUE}
library(knitr)
opts_chunk$set(echo=TRUE, cache=F)
```

## Loading and preprocessing the data

- download the data from the webpage:
```{r, download}
# set the wd, in my case it is:
setwd("~/Desktop/DataScience/ReproducibleResearch/project1/RepData_PeerAssessment1")

# download data from the website if it is not loaded
if(!file.exists("activity.zip")) {
        file.Url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(file.Url, destfile="activity.zip", method="curl")
}
```

- unzip and read the data
```{r read and load}
unzip("activity.zip")
activity <- read.csv("activity.csv", stringsAsFactors=FALSE, na.strings="NA")
```

- transform the date to dateformat
```{r transform}
library(lubridate)
activity$date <- parse_date_time(activity$date, "ymd")
```

## What is mean total number of steps taken per day?

- calculate the total number of steps taken per day ignoring missing data
```{r }
library(dplyr)
# take the sum per day ignoring NA's
sumactivity <- activity %>% filter(steps != is.na(steps)) %>% group_by(date) %>% summarise(number = n(), total_steps = sum(steps))
```

- make a histogram of the total number of steps taken each day
```{r histogram}

library(ggplot2)

# plot a histogram
total <- ggplot(sumactivity, aes(x=total_steps))
plottotal <- total + geom_histogram(binwidth=1000, colour="black", aes(fill = ..count..)) + scale_fill_gradient("Count", low = "#CCCCCC", high = "#333333") + ggtitle("Histogram of total steps per day") + xlab(NULL)
plottotal
```

- calculate and report mean and median ignoring missing values
````{r mean and median}
mean <- mean(sumactivity$total_steps, na.rm=TRUE)
mean
roundmean <- as.integer(mean)

median <- median(sumactivity$total_steps, na.rm=TRUE)
median
```

## What is the average daily activity pattern?

- make a time series plot (i.e. type = "l", the ggplot equivalent is geom_line()) of the 5 minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r time series1}
activity$interval <- as.factor(activity$interval)
sumactivity <- activity %>% filter(steps != is.na(steps)) %>% group_by(interval) %>% summarise(number = n(), steps = mean(steps))

interval <- ggplot(sumactivity, aes(x=interval, y=steps, group=1))
interval + geom_line(colour="blue") + scale_x_discrete(breaks= seq(0, 2000, by=1000)) + ylab("average number of steps") + ggtitle("Average of all steps of the 5 minute interval across all days")
```

- find the 5-minute interval that, on average, contains the maximum number of steps.
````{r }

maxinterval <- sumactivity %>% filter(steps == max(steps)) %>% select(interval)

```
The 5-minute interval that, on average, contains the maximum number of steps is `r maxinterval`

## Imputing missing values

- calculate and report the total number of missing values in the dataset
```{r missing data}
missing_values <- nrow(activity[is.na(activity$steps),])
````
The total number of missing values is `r missing_values`.


### Strategy for including missing values
- split the data into to groups one without missing values and one with missing values
```{r filling missing data}

avaliable <- filter(activity, !is.na(steps))
missing <- filter(activity, is.na(steps))
```

- fill the missing values with a random value from the set without missing values matching the same interval
```{r filling missing data2}
set.seed(100) # always set the seed for reproducibility
for(i in 1:nrow(missing)){
        missing$steps[i] <- sample(avaliable$steps[avaliable$interval == missing$interval[i]],1)
}
```

- merge the two data sets together
```{r }
newdata <- data.frame(rbind(avaliable,missing))
```

- order them by date and interval to get the original order back
```{r }
newdata <- arrange(newdata, date, interval)
```

- calculate the total number of steps taken per day
```{r }
summissing <- newdata  %>% group_by(date) %>% summarise(number = n(), total_steps = sum(steps))
```

- make a histogram of the total number of steps taken each day 
```{r Histogram of total steps per date with the missing data filled in}
totalmissing <- ggplot(summissing, aes(x=total_steps))
plotmissing <- totalmissing + geom_histogram(binwidth=1000, colour="black", aes(fill = ..count..)) + scale_fill_gradient("Count", low = "#CCCCCC", high = "#333333") + ggtitle("Histogram of total steps per day with the missing data filled in") + xlab(NULL)
plotmissing
```

- calculate and report the mean and median total number of steps taken per day
```{r }
mean_missing <- mean(summissing$total_steps)
mean_missing
roundmean_missing <- as.integer(mean_missing)
median_missing <- median(summissing$total_steps)
median_missing
```
The mean including the missing values is with `r roundmean_missing`  smaller than the mean without missing values, which is `r roundmean`
The median including the missing values ist `r median_missing`. and doesn't change much compared to the median without the missing values, which is `r median`.


## Are there differences in activity patterns between weekdays and weekends?
- create a new factor variable in the dataset with two levels(“weekday” and “weekend”) indicating whether a given date is a weekday or weekend day.
```{r }
library(lubridate)
newdata <- newdata %>% mutate(daysofweek = wday(date)) %>% mutate(days = ifelse(daysofweek > 1 & daysofweek < 7, "weekday", "weekend"))

# convert it to a factor variable
newdata$days <- factor(newdata$days)
```

- make a panel plot containing a time series plot (i.e. type = "l", the ggplot2 equivalent is geom_line()) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r times series2}
sumnewdata <- newdata %>% group_by(interval,days) %>% summarise(average_steps = mean(steps))

# reorder the levels
sumnewdata$days <- relevel(sumnewdata$days, ref = "weekend")

timeseries <- ggplot(sumnewdata, aes(y=average_steps, x=interval, group=1))
timeseries + geom_line(colour="blue") + facet_wrap(~days, ncol=1) + scale_x_discrete(breaks=c(0,500,1000,1500,2000)) + ylab("average number of steps") + ggtitle("Average of all steps of the 5 minute interval across all days")
````

During weekdays more steps are recorded in the earlier intervals and the average steps peaks higher, while on weekend more average steps are recorded in the later intervals.